### 2009-12-01 docs/marcello-dissertation/thesis/chapter6-data-persistence-component-design-and-architecture.tex
Updates to the images of the experiments and revisions to different chapters from 4 to 10.
git-svn-id: svn://localhost/netbeams/branches/marcello/persistence@664 559edde2-a4d4-44e0-96fc-4035e40d7396

\section{Business Process Analysis}
\label{sec:business-process-analysis}
As described in section \ref{chap:netbeams-overview}, the DSP Platform was
designed to support an environmental sensor network based in San Francisco
Bay, composed by different types of sensors. The process of extracting the
measurements from the sensors to be reused using the DSP Platform can be
summarized following the UML Business Process Diagram \cite{uml} depicted in
figure \ref{fig:dsp-persistence-business-process}. It is important to note
that this diagrams details remote communication between sensor devices and a
centralized server, as part of the main scenario to solve the problem of data
persistence.

@@

\subsection{Functional Requirements}
As part of the user stories related to data access, figure
\ref{fig:DSP-Data-Persistence-UseCases-Diagram-Users} describes the use cases
that the end users can perform with the system.

@@

\begin{figure}[!b]
  \centering
  \includegraphics[scale=0.5]{../diagrams/DSP-Data-Persistence-UseCases-Diagram-System}
  \caption{UML Use Case diagram for System Persistence Functions}
  \label{fig:DSP-Data-Persistence-UseCases-Diagram-System}
\end{figure}

@@

\begin{itemize}
  \item The DSP Data Persistence component must maintain the received DSP
  Measurement Messages in-memory for a specific rate in order to decrease the
  write load on the database;
  \item The Persistence storage system must be able to cope with dozens of DSP
  Messages;
  \item The Persistence storage system must be able to save measurements quickly;
  \item The Persistence storage system must be able to scale without too much
  architectural changes;
  \item The Persistence storage system must not impose the knowledge of
  complicated database system languages such as SQL \cite{sql} or XML XPath
  \cite{xml-xpath}, but use access through API calls, since it is closer to
  researchers.
\end{itemize}

\section{Data Model Design}
\label{sec:dsp-persistence-data-model}
As described in the previous chapter, the data model that might give better
support for collected data from sensor networks is the Key-Value-Pair Data
Model. Beyond the data collected from the sensor devices, Data Provenance 
techniques must also be considered to better describe the collected data. In
general, the data model must include the following metadata besed on Data
Provenance:

\begin{itemize}
  \item \textbf{Locality}: properties that describe the location of the data.
  For example, the host machine name or IP address, the GPS coordinate
  \cite{gps};
  \item \textbf{Identity}: what uniquely identities the data among the others
  produced by other sensor devices in the network, as for instance a
  UUID\footnote{Universally Unique Identifier};
  \item \textbf{Time-Space}: when the data was ``sensed'' \cite{db-provenance}
  and when the collection transaction happened \cite{sn-time-series}.
\end{itemize}

@@

\begin{itemize}
  \item \textbf{sensor\underline{ }ip\underline{ }address}: it's extracted from
  the DSP Message Produce and identifies which sensor generated the sampling;
  \item \textbf{message\underline{ }id}: it's extracted from each of the messages contained
  in the DSP Message Container;
  \item \textbf{transaction\underline{ }time}: it's extracted from the DSP message container
  creation time and it is used to identify when the transaction occurred;
  \item \textbf{fact\underline{ }time}: it's extracted from the SondeDataContainer's date
  and time, and identifies the time in which the collected data occurred;
  \item \textbf{latitude}: the latitude coordinate of the position of
  the sensor device, captured from references of the device (not supported by
  NetBEAMS);
  \item \textbf{longitude}: the longitude coordinate of the position of the
  sensor device, captured from references of the device (not supported by
  NetBEAMS).
\end{itemize}

@@

\begin{itemize}
  \item data: this key defines the values of the document, and will have every
  different property of the sensor. For instance, the data key for the
  properties of the YSI device will contain the set of properties defined for
  the device, such as battery, temperature, salinity, etc.
\end{itemize}

@@

The new proposed Data Persistence component must follow the DSP requirements
for a new DSP Component following the the design-pattern of Data
Producer/Consumer. In a nutshell, the design of a DSP Component resulting
maintains the dependency among 3 different major Java packages, as the addition
of such a DSP Data Persistence component depends on 2 existing packages, namely
the OSGi Framework and the DSP Platform, as well as and an additional package
for the database access driver API, as shown in the UML Package Dependency
Diagram \cite{uml} of figure
\ref{fig:DSP-Data-Persistence-Packages-Dependency}.

@@

The DSP Data Persistence component starts with the design of an OSGi Activator
class, and the implementation of the DSP Component, resposible to provide
services to the platform. Based on these specifications, the following set of
classes in figure \ref{fig:DSP-DataPersistence-Activator-Class-Diagram} shows
the UML Class Diagram \cite{uml} implemented for the component.

@@

The behavior of the activator class follows the specifications of the
OSGi Framework, and for this reason, instances of the DSP Data Persistence
Activator can be in the states described on the UML State Diagram depicted in
figure \ref{fig:DSPPlatform-Install-Usage-State-Diagram}, as it is an extension
of the OSGI class BundleActivator, as well as being responsible for stopping
or starting the the DSP Data Persistence component instance. Following the
specifications of section \ref{netbeams-architecture}, the DSP Data Persistence
Platform can be initialized as described by the UML Sequence Diagram of figure
\ref{fig:From-OSGi-Framework-to-DSP-Data-PersistenceActivator-Sequence-Diagram}:

@@

Any configuration parameter required by a DSP Component is sent during its
bootstrap process through the use of an optional DSP Update Message, as
described in section \ref{sec:dsp-message}. The initialization of the DSP
Component Activator and the DSP Component registration are also shown in
figure \ref{fig:From-OSGi-Framework-to-DSP-Data-PersistenceActivator-Sequence-Diagram}.

@@

The the gate "Start DSP Activator Gate" reaches the DSP Data Persistence, it
first creates the unique instance of the class DSP Data Persisence and uses
that instance to register it as the OSGi service through the call to the DSP
Platform helper class ActivatorHelper. At this point, the DSP Data Persistence
has been installed and is in the state "Active".

@@

The DSP Data Flusher is a thread that is designed to be in two different
states as depicted in figure
\ref{fig:DSP-DataPersistence-Flusher-State-Diagram}: RUNNABLE and
TIMED\underline{ }WAITING. In the former, the DSP Data Flusher will be
contacting the Transient Persistence Layer to verify if there are any message
waiting to be sent to the Database service, while in the latter the DSP Data
flusher will be waiting for the next cycle defined by the bootstrap parameter
TRANSIENT\underline{ }DATA\underline{ }FLUSHER\underline{ }DELAY.

@@

\begin{itemize}
  \item A bootstrap DSP Update Message is delivered to the DSP Data Persistence
  component when the DSP Platform is initializing the component. In this way,
  the DSP Data Persistence can initialize the DspDataFlusher worker thread to
  flush temporary messages into the database at the rate of
  TRANSIENT\underline{ }DATA\underline{ }FLUSHER\underline{ }DELAY in seconds;
  \item Any DSP Measure Message can be persisted by the DSP Data Persistence
  component. When the DSP Broker delivers a DSP MeasureMessage to this
  component, it will add the message to a temporary memory location to decrease
  I/O on the database, as depicted by the gate "create PersistentMessageUnit
  Gate". The simple procedure to add the message into the transient persistent
  layer.
\end{itemize}

The class PersistenceMessageUnit is the major transient persistence unit that
carries references regarding the originating DSP Message that was collected by
the DSP Data Persistence component. It is composed by a SensorLocation and
contains a reference to the PersistentMessageState enumaration. The former
identifies which sensor produced the collected data from the DSP Message by the
IP address, while the latter identifies whether the PersistenceMessageUnit has
been saved into the Database or not as depicted by the UML State Diagram
\cite{uml} in figure \ref{fig:PersistentMessageState-Diagram}.

@@

According to the data model specifications defined in section
\ref{sec:dsp-persistence-data-model}, the persistence model must carry
information regarding which sensor device produced the collected data. For
this reason, an instance of the class SensorLocation carries the information
about the IP address from the component that originally produced the DSP
Message. Then, an instance of the class PersistenceMessageUnit is created,
having its state on TRANSIENT, as detailed on UML State Diagram \cite{uml} in
figure \ref{fig:PersistentMessageState-Diagram}.

>>>>

\section{Business Process Analysis}
\label{sec:business-process-analysis}
As described in section \ref{chap:netbeams-overview}, the DSP Platform was
designed to support an environmental sensor network based in San Francisco
Bay, composed by different types of sensors. The process of extracting the
measurements from the sensors to be reused through the DSP Platform can be
summarized following the UML Business Process Diagram \cite{uml} depicted in
figure \ref{fig:dsp-persistence-business-process}. It is important to note
that this diagrams details remote communication between sensor devices and a
centralized server, as part of the main scenario to solve the problem of data
persistence.

@@

\subsection{Functional Requirements}
\label{sec:use-cases}

As part of the user stories related to data access, figure
\ref{fig:DSP-Data-Persistence-UseCases-Diagram-Users} describes the use cases
that the end users can perform with the system.

@@

\begin{itemize}
  \item \textbf{Search Measurements by Sensor Properties}: users of the
  collected data must be able to search data through the properties of the
  sensors. This is related to identity principle of what was collected
  as provenance \cite{db-provenance};
  \item \textbf{Search Measurements by Sensor Location}: users of the collected
  data must be able to search data by a given sensor location, as referred to
  the provenance from where the data was collected;
  \item \textbf{Search Measurements by Data collection date}: users of the
  collected data must be able to search data by a given date, or ranges of
  start and end date;
  \item \textbf{Access the data through API, Programming Languages}: users of the
  collected data must be able to search data by the use of different
  programming and scripting languages of their preference;
  \item \textbf{Export Data to OPEnDAP format}: users of the collected data
  must be able to export the collected data to the OPEnDAP format;
  \item \textbf{Export Data to Spreadsheets}: users of the collected data must
  be able to export the collected data to spreadsheets.
\end{itemize}

@@

\begin{itemize}
  \item The DSP Data Persistence component must maintain the received DSP
  Measurement Messages in-memory for a specific rate in order to decrease the
  write load on the database;
  \item The Persistence storage system must be able to cope with dozens of DSP
  Messages;
  \item The Persistence storage system must be able to save measurements quickly;
  \item The Persistence storage system must be able to scale without too much
  architectural changes, or schema changes;
  \item The Persistence storage system must not impose the knowledge of
  complicated database system languages such as SQL \cite{sql} or XML XPath
  \cite{xml-xpath}, but use access through API calls, as it offers better
  abstractions for non-computer science researchers
  \cite{sn-programming-language}.
\end{itemize}

\section{Data Model Design}
\label{sec:dsp-persistence-data-model}

As described in the previous chapter, the data model that might give better
support for collected data from sensor networks is the Key-Value-Pair Data
Model. Beyond the data collected from the sensor devices, Data Provenance 
techniques must also be considered to better describe the collected data
\cite{sn-provenance}. In general, the data model must include the following
metadata besed on Data Provenance:

\begin{itemize}
  \item \textbf{Locality}: properties that describe the location of the data.
  For example, the host machine name or IP address, the GPS coordinate
  \cite{gps};
  \item \textbf{Identity}: what uniquely identities the data among the others
  produced by other sensor devices in the network, as for instance a
  UUID\footnote{Universally Unique Identifier};
  \item \textbf{Time-Dimensions}: when the data was ``sensed''
  \cite{db-provenance} and when the collection transaction happened
  \cite{sn-time-series}.
\end{itemize}

@@

\begin{itemize}
  \item \textbf{sensor.ip\underline{ }address}: it is extracted
  from the DSP Message Produce and identifies which sensor generated the sampling;
  \item \textbf{sensor.location.latitude, longitude}: the latitude and
  longitude GPS coordinates of the position of the sensor device, captured from
  references of the device. This is currently provided as a parameter in the
  bootstrap message for the component;
  \item \textbf{message\underline{ }id}: it's extracted from each of the messages contained
  in the DSP Message Container;
  \item \textbf{time.transaction}: it is extracted from the DSP
  message container creation time and it is used to identify when the transaction occurred;
  \item \textbf{time.fact}: it is extracted from the
  SondeDataContainer's date and time, and identifies the time in which the
  collected data occurred.
\end{itemize}

@@

\begin{itemize}
  \item \textbf{observation}: this key defines the values of the document, and
  will have every different property of the sensor. For instance, the data key for the
  properties of the YSI device will contain the set of properties defined for
  the device, such as battery, temperature, salinity, etc.
\end{itemize}

@@

The new proposed Data Persistence component must follow the DSP requirements
for a new DSP Component, which specifies the design-pattern of Data
Producer/Consumer. In a nutshell, the design of a DSP Component maintains the
dependency among 3 different major Java packages: it depends on 2 existing
packages, namely the OSGi Framework and the DSP Platform, as well as and an
additional package for the database access driver API, as shown in the UML
Package Dependency Diagram \cite{uml} of Figure
\ref{fig:DSP-Data-Persistence-Packages-Dependency}.

@@

The DSP Data Persistence component starts with an OSGi Activator class, and the
implementation of the DSP Component, resposible to provide services to the
platform. Based on these specifications, the following set of classes in
figure \ref{fig:DSP-DataPersistence-Activator-Class-Diagram} shows the UML
Class Diagram \cite{uml} implemented for the component.

@@

The behavior of the activator class follows the specifications of the
OSGi Framework and, for this reason, instances of the DSP Data Persistence
Activator can be in the states described on the UML State Diagram depicted in
Figure \ref{fig:DSPPlatform-Install-Usage-State-Diagram}. It is important to
note that the activator is an extension of the OSGI class BundleActivator, as
well as being responsible for stopping or starting the the DSP Data Persistence
component instance. In this way, following the specifications of section
\ref{netbeams-architecture}, the DSP Data Persistence Platform can be
initialized as described as it is described by the UML Sequence Diagram of
Figure
\ref{fig:From-OSGi-Framework-to-DSP-Data-PersistenceActivator-Sequence-Diagram}:

@@

Any configuration parameter required by a DSP Component is sent during its
bootstrap process through the use of an optional DSP Update Message, as shown
in Listing \ref{file:dsp-data-persistence-bootstrap.xml}. The initialization of
the DSP Component Activator and the DSP Component registration are also shown
in figure \ref{fig:From-OSGi-Framework-to-DSP-Data-PersistenceActivator-Sequence-Diagram}.

@@

As soon as the gate "Start DSP Activator Gate" reaches the DSP Data
Persistence instance, it first creates the unique instance of the class DSP
Data Persisence and uses that instance to register it as the OSGi service
through the call to the DSP Platform helper class ActivatorHelper. At this
point, the DSP Data Persistence has been installed and is in the state
"Active".

@@

The DSP Data Flusher is a thread that is designed to be in two different
states as depicted in figure
\ref{fig:DSP-DataPersistence-Flusher-State-Diagram}: RUNNABLE and
TIMED\underline{ }WAITING. In the former state, the DSP Data Flusher will be
contacting the Transient Persistence Layer in order to verify if there are any
message in the outbound queue ready to be transmitted sent to the Database
service, while the DSP Data flusher will be waiting for the next cycle defined
by the bootstrap parameter TRANSIENT\underline{ }DATA\underline{
}FLUSHER\underline{ }DELAY in the latter state.

@@

\begin{itemize}
  \item A bootstrap DSP Update Message is delivered to the DSP Data Persistence
  component when the DSP Platform is initializing the component. In this way,
  the DSP Data Persistence can initialize the DspDataFlusher worker thread to
  flush temporary messages into the database at the rate of
  TRANSIENT\underline{ }DATA\underline{ }FLUSHER\underline{ }DELAY in seconds,
  among others. Listing \ref{file:dsp-data-persistence-bootstrap.xml} shows an
  implementation of a bootstrap message for the DSP Data Persistence;
  \item Any DSP Measure Message can be persisted by the DSP Data Persistence
  component. When the DSP Broker delivers a DSP MeasureMessage to this
  component, it will add the message to a temporary memory location to decrease
  I/O on the database, as depicted by the gate "create PersistentMessageUnit
  Gate". This last gate describes the simple procedure to add the message
  into the transient persistent layer. Listing
  \ref{file:dsp-message-serialized-ysi}.
\end{itemize}

The class PersistenceMessageUnit is the major transient persistence unit that
carries references regarding the originating DSP Message that was collected by
the DSP Data Persistence component. It is composed by a SensorLocation and
contains a reference to the PersistentMessageState enumaration. The former
identifies which sensor produced the collected data from the DSP Message by the
IP address, while the latter identifies whether the PersistenceMessageUnit has
been saved into the Database or not as depicted by the UML State Diagram
\cite{uml} in Figure \ref{fig:PersistentMessageState-Diagram}.

@@

According to the data model specifications defined in section
\ref{sec:dsp-persistence-data-model}, the persistence model must carry
information regarding which sensor device produced the collected data. For
this reason, an instance of the class SensorLocation carries the information
about the IP address and and optional GPS location from the component that
originally produced the DSP Message. Then, an instance of the class
PersistenceMessageUnit is created at the state ``TRANSIENT'', as detailed
on UML State Diagram \cite{uml} in Figure \ref{fig:PersistentMessageState-Diagram}.

### 2009-12-01 docs/marcello-dissertation/thesis/chapter8-experimental-results-correct-behavior-and-performance.tex
Updates to the images of the experiments and revisions to different chapters from 4 to 10.
git-svn-id: svn://localhost/netbeams/branches/marcello/persistence@664 559edde2-a4d4-44e0-96fc-4035e40d7396

\subsection{Key-Value definition}

In document-oriented databases, there is no definition of table. In constrast,
the entities are persisted using any data arrangment, in any order, attributes
and the associated values. The definition of the document for mongoDB takes
into account the properties defined by the data provenance study defined by
\cite{sn-provenance}. The following list are the definitions of the values of
the defined keys and the provenance-aware document, as an example described by
Listing \ref{file:mongodb-ysi-data-format}:

@@

\subsection{Workload}

@@

In order to verify the system capacity to scale, the data will be devided into
different shards as part of the support to distribute system, that is, in a
Data-Centric way.

\begin{itemize}
  \item First Round: 483,840 items worth one year of produced YSI data at the
  rate of 1 observation per minute;
  \item 483,840 * 5, or 2,419,200, representing 5 different YSI components
  producing data every minute;
\end{itemize}

@@

The definition for the single server follows a regular specification of the
mongoDB database system. All the collected data from sensors are stored in a
given Single Server, which characterizes an External Storage device.

@@

\subsection{Scenarios}

In order to verify the feasibility of the system, the use of different search
scenarios are used:

@@

\section{Measurements Results}

The result of the insertion of data in different scenarios are shown in the
different logs captured during the experiments. First, the avarage insertion
times gives the capacity of the system with and without the definition of
indexes over the data. Then, the numbers regarding the scenarios described
earlier will be shown:

@@

The insersion time avarages are regarding the different approaches to model a
document in the database system:

\begin{itemize}
  \item No Index : Insert Avarage = xyz 
  \item Indexed : Insert Avarage = xyz
\end{itemize}

On the other hand, the avarages for each of the scenarios are devided by the
amount of data produced by one single device or five different devices.

For one single device, the values can be summarized as follows:
\begin{itemize}
  \item Not Index : A) = xyz
  \item Indexed : A) = xzy
  \item Not Index : B) = xyz
  \item Indexed : B) = xzy
  \item Not Index : C) = xyz
  \item Indexed : C) = xzy
\end{itemize}

For data worth five different devices, the values can be summarized as follows:
\begin{itemize}
  \item Not Index : A) = xyz
  \item Indexed : A) = xzy
  \item Not Index : B) = xyz
  \item Indexed : B) = xzy
  \item Not Index : C) = xyz
  \item Indexed : C) = xzy
\end{itemize}

\subsection{Data-Centric Cluster Environment}

The insertions in a sharded environment are added the operation of finding
which ``bucket'' the data must be placed. mongoDB gives support for automatic
sharding after the definition of a key based on range of the values. The
avarages of the insertions can be shown as follows:

\begin{itemize}
  \item NO Index : Insert Avarage = xyz
  \item Indexed : Insert Avarage = xyz
\end{itemize}

For one single device, the values can be summarized as follows:

\begin{itemize}
  \item Not Index : A) = xyz
  \item Indexed : A) = xzy
  \item Not Index : B) = xyz
  \item Indexed : B) = xzy
  \item Not Index : C) = xyz
  \item Indexed : C) = xzy
\end{itemize}

For data worth five different devices, the values can be summarized as follows:

\begin{itemize}
  \item Not Index : A) = xyz
  \item Indexed : A) = xzy
  \item Not Index : B) = xyz
  \item Indexed : B) = xzy
  \item Not Index : C) = xyz
  \item Indexed : C) = xzy
\end{itemize}

@@

mongoDB is that the data can be distributed using Horizontal Partitioning,
which mitigates problems related to referencial integrity.

\begin{itemize}
  \item Strengths as compared to literature
  \item Weaknesses as compared to literature
\end{itemize}

Data-Centric Storage and Database Shards = Horizontal scalability for data
based on specific keys = Scalability facing High Volume of Collected Data

Weakness, Technological, and skills-based.

>>>>

\subsection{Key-Value definition}
\label{sec:keys-definition}

In document-oriented databases, there is no definition of table. In constrast,
the entities are persisted using any data arrangment, in any order, attributes
and the associated values, as shown in Listing
\ref{file:mongodb-ysi-data-format}. The definition of the document for mongoDB
takes into account the pairs of key and values related for the collected data
from the sensors, as well as the properties defined by the data provenance
study \cite{sn-provenance}. The collected data is stored in a key called
``observation''. On the other hand, the provenance-aware keys are defined as
follows:

@@

\subsection{Workload}
\label{sec:workload}

@@

\begin{itemize}
  \item \textbf{First Round}: 483,840 items worth one year of produced YSI data
  at the rate of 1 observation per minute;
  \item \textbf{Next Rounds}: 483,840 * 5, or 2,419,200, representing 5
  different YSI components producing data every minute.
\end{itemize}

@@

The definition for the single server follows a regular specification of the
mongoDB database system. All the collected data from sensors are stored in a
given single server, which characterizes an External Storage device.

@@

\subsection{Scenarios}
\label{sec:exp-scenarios}

In order to verify the feasibility of the system, the use of different search
scenarios are used defined based on the use cases of section
\ref{sec:use-cases}:

@@

Moreover, the experiments must also export data in different formats such as 
spreadsheets and use different programming languages APIs to access the data.

\section{Measurements Results}
\label{sec:exp-measurements}

The result of the insertion of data in different scenarios are shown in the
different logs captured during the experiments. First, the avarage insertion
times gives the capacity of the system with and without the definition of
indexes over the data. Then, the numbers regarding the scenarios described
earlier will be shown.

@@

The insersion time avarages an impressive ~169,508 instances per minute during
the experiments using the Java driver to connect with the mongoDB server. The
following lists the storage consumed by the generated data described in the
section \ref{sec:workload}.

\begin{itemize}
  \item 1 YSI worth of data = 234.16 MB 
  \item 5 YSI worth of data = 1.23 GB
\end{itemize}

On the other hand, the avarages for each of the scenarios are devided by the
amount of data produced by one single device or five different devices. For
data worth five different devices, without indexes in the database, the values
can be summarized as follows:

\begin{itemize}
  \item A) = ``time.valid'' for week period, or 7 days = 3820ms
  \item B) = a given IP address by ``sensor.ip\underline{ }address'' = 3721ms
  \item C) = given values for the keys ``observation.Salinity'' and 
  ``observation.WaterTemperature'' = 3167ms
\end{itemize}

@@

The implementation of a data persistence for any sensor network must take into
the characteristics of the sensor network, the location of the collected data,
etc. As described in chapter 4, the taxonomies classify different aspects of
the persistence of data. Along with it, the design and implementation of a data
persistence layer for NetBEAMS must meet the requirements and specifications
described in chapter 5 and chapter 6, respectively. This section discusses the
experiments scenarios described to evaluate a real-world experiment relating
the implementation and results to each of the taxonomies. Furthermore, the
experiments provided by this work cover a one year worth of data for one and
five YSI devices and evaluates the use cases defined in section
\ref{sec:exp-scenarios}.

\subsection{Taxonomic Evaluation}

Any database system covers the purpose of the sensor data for NetBEAMS, which
is Data Archival, along with the type of storage for the data as an External
Storage. After selecting mongoDB, the development of the persistence layer met
the requirements of archiving data after the randomly generated data for the
YSI sonde was persisted. As it is shown in section \ref{sec:exp-measurements},
the collected data is archived in files using the BSON format \cite{bson}, the
binary representation used to store instances of the document described in
Listing \ref{file:mongodb-ysi-data-format}. Along with the purpose of data
archival, the use of mongoDB also met the specification of an external storage
since the data persisted in the file system.

One of the requirements for NetBEAMS is the use of a schema-less data model.
mongoDB uses a variation of the key-value data representation based on the JSON
\cite{json}, and provides APIs in different languages for the data creation
process. In this way, the implementation of the persistence model captures the
instances of any known data collected from NetBEAMS and saves it using the
document pattern as shown in Listing \ref{file:mongodb-ysi-data-format}. One
application of a schema-less data model is the addition of any new key-value
pair to any instance of the document, as mongoDB supports update operations to
any of the collections. For instance, annotation in sensor networks is
widely used to describe the findings of a particular data set
\cite{sn-annotation-ex1, sn-annotation-ex2}. In case of an oil spill in the San
Francisco Bay occurs, researchers can update the data set of a given date range
by adding tags to the relating documents. As a consequence of the use of a
schema-less data model, these changes do not affect the other data instances,
since there are no schema changes.

As described in chapter 5, the schema-less data model can provide a better
support to Data Provenance because the data model gives support to the change
of records without affecting the others. In this way, the metadata used by the
implementation described in the previous chapter clearly describes the three
fundamental questions regarding the collected data from sensor devices, as
described in section \ref{sec:keys-definition}. First, the identity of the data
is given by the key ``message\underline{ }id'', in a way that it uniquely
identifies the data produced by the sensor identified by the key
``sensor.ip\underline{ }address''. In this way, the message can be tracked back
to the sensor that produced the data for different purposes such as tracking
the data producer. Second, time dimensions were used in order to track two
different aspects of the data collection: the valid time can be used to
correlate observations to the exact moment it occurred, while the transaction
time can be used to verify when the data was collected, and make decisions such
as deleting the instances of data were collected in the previous two days, etc.
Last, but not least, the use of the key ``sensor.location'' can give the exact
location of the data, since researchers can be interested in finding data based
on a specific location. Finally, the most important aspect of the schema-less
data model is that the observed data is added to the key ``observation'', as
any sensor device carries its set of attributes of key-value pairs. A good
example of the applicability of the schema-less data model is the scalability
regarding changes to the data structure. For instance, if an existing sensor
device is flushed with a new version of the firmware\footnote{The first set of
machine instructions to run on the hardware after the application of power stored
permanently in PROM or ROM or semi-permanently in EPROM}, changing the format
of data types or adding new values, the following instances of collected data
may be different to the previous one by simply adding the new additions of
data. For this reason, the DSP Data Persistence gives administrators the
ability to choose which fields to persist by declaring it using a bootstrap
message for the component, as shown in the parameter
``YSI\underline{ }DESIRED\underline{ }PROPERTIES'' of Listing
\ref{file:dsp-data-persistence-bootstrap.xml}.

The centralized query processing used by NetBEAMS is a result of the
infrastructure used by NetBEAMS, as well as the purpose of data and the
location of the collected data. As a direct result of a schema-less data
model, the centralized query processing is easily developed using any of the
available programming languages provided by mongoDB. Considering the fact that
programming languages are easier abstractions to researchers without computer
science background \cite{sn-programming-language}, mongoDB's approach to data
access through API can be seen as an easier task to manage data. As it is shown
in Listing \ref{file:experiment-query-scenarios}, the scenarios designed to
evaluate the implementation are written in Javascript, since it is the script
language by the mongoDB Shell. Therefore, it attends the requirement of not
using the SQL language as the main query language of the data.

One of the most important features of a data persistence for sensor
networks is the replacement of the SQL by the use of programming
languages, as they tend to be easier abstraction to users without the
background in database systems \cite{sn-programming-language}. In this way,
mongoDB covers the requirements of providing data access to the collected
sensor data in different languages. For example, the data collected by the DSP
Data Persistence component is written in Java, while the implementation of the
experiments scenarios for querying the collected data is given in Javascript. In
this way, researchers of the RTC may choose one of the enumarated languages
APIs provided by the mongoDB to access the stored data from NetBEAMS.

Finally, the implementation supported by mongoDB is either in a single or
distributed servers. The former approach is done as a regular database server
on a single host machine, where its main goal is to manage the collected sensor
data. On the other hand, however, the latter approach can be used to improve
the performance of the data selection in larger data sets by offering different
capabilities such as Database Shards and Data Replication.

\subsection{General Evaluation}

In general, the implementation of the DSP Data Persistence component meets
the specification of the use cases defined in section \ref{sec:exp-scenarios}.
The implementation of mongoDB using a single server environment can give
support to the current workload and infrastructure from the use of NetBEAMS to
collect data from the sensors. From the perspective of disk-space utilization
and given that ~1.5 Gigabyte of disk storage is necessary to store
one-year-worth of collected data for five YSI devices, the implementation
scales to 80 years of data in a server with 120 Gb of available disk space.
However, if the database server runs out of space, the addition of a new
database shard can be done.

The performance to insert data into the database system using the Java driver
is documented by mongoDB as the fastest approach, providing persistence to
approximaly 169,508 instances of data. Considering this number of insertions to
the database, it is clear to say the mongoDB implementation can virtually
support around 169 thousands of sensors a minute. Similarly, the implementation
of a data-centric scenario using mongoDB can improve the performance of the
system, given that the schema-less data model gives the ability to partition
data horizontaly. In this way, latency related to data constraints and data
integrity checks are avoid because an entire instance of the collected data
resides in the same ``chunck'' of data. This is the main advantage over the
relational model, previously discussed in chapter 5.

One possibility of improving performance on the given implementation is the use
of indexes on each of the keys. As a direct consequence of this procedure,
insertion time to this approach is approximaly decreased by a 10th of the
insertion time experienced with the experiments without indexes. Similarly, it
turns out that the use of indexes improves the query processing by 5 times when
indexes are not used. Therefore, the use of indexes should be enforced in a
real-world deployment, since both requirements for insertion and retrieval are
met.

When it comes to re-using the data collected by the DSP Data Persistence,
different approaches are available. First, the use of the export capabilities
to a common-delimited values was performed \ref{file:mongodb-export-command},
taking approximaly 3.3 minutes to export one year worth of data for 5 YSI
devices. Second, the use of the client shell gives a centralized access to
either the single server of sharded server option. However, in order to better
provide access of data to users without skills in database systems, web
applications can be built on top of the REST \cite{http-rest} Web Services APIs
provided by mongoDB. One such example of using the REST API is the adaptation
of Apache Futon 4 mongoDB, as shown in Figure
\ref{fig:view-collections-instance-browser-futondb}, where users can navigate
through the collections of the collected sensor network data categorized by the
collection. Therefore, RTC staff technicians could take advantage of this tool
to view the collected data.

The design and implementation of the guidelines designed by the taxonomies can
be applied to any sensor network that produces data with the purpose of Data
Archival in External or Data-Centric storage.

### 2009-12-01 docs/marcello-dissertation/thesis/chapter9-conclusions-and-future-works.tex
Updates to the images of the experiments and revisions to different chapters from 4 to 10.
git-svn-id: svn://localhost/netbeams/branches/marcello/persistence@664 559edde2-a4d4-44e0-96fc-4035e40d7396

This work presented a solution to store data from sensor networks.....
...the main result of this work was that the persistence is

There are many important features covered in this work, including...
  - Use of open-source UI for KVP Sensor Data

Other important features that COULD HAVE BEEN covered, includes\ldots
  - Different Dababase system, use of cloud computing, etc
  - Use of MapReduce technology and implementations, distributed Processing that scales the proposed KVP storage

... The main results from this dissertation were important because of the contributions on new technologies...
  - An alternative Data Model for Data Storage in Sensor Networks
  - An alternative apprach for Data-Centric Storage in Sensor Networks

\section{TODO list}

>>>>

This work presented a collection of taxonomies related to data persistence for
sensor networks using different approaches ranging from data models and server
infrastructure organization. The main result of this work was that persistence
for collected data from sensors are better maintained using schema-less data
models, since it provides offers an approach to deal with unanticipated data
types from the sensor devices.

There are many important features covered in this work, including the
importance of the sensor networks rely in how users have access to the
collected data. In order to provide access to the collected data, one must
first assess the different types of infrastructural characteristics of the
studied sensor network. In general, the selection of a technology that
provides such persistence capabilities can be part of a process of analysis of
taxonomies proposed by this work. However, this work proposes a data model not
yet used in the sensor network community and, thus, represents an important
contribution for the area.

Taking into account the software architecture provided by NetBEAMS, as well as
the requirements for data collection from the RTC research group, this work
shows a novice way to persist data collected from sensor devices by using a
schema-less data model, whose query capabilities are based upon programming
languages, in contrast to the use of the Structured Query Language, SQL. The
experiment results showed that the implementation of a persistence layer with
mongoDB can be used to persist data at its maximum rate of one sample per
minute. However, performance can be improved by using techniques such as
database sharding, since it offers a direct implementation capability to
Data-Centric Storage approach. This approach decreases the amount of time to
search data since it distributes the data instances across different cluster
nodes.

There are many different directions for future works. First, the Data-Centric
implementation performance may be extremely improved by using the technique
called MapReduce \cite{map-reduce}. This approach allows mongoDB to search
different machines in parallel whenever the data requested is located in
different database shard hosts. In addition to that, the use techniques to
cluster collected data before they are inserted into the database may be
important as shown in \cite{sn-time-series}. One of the key issues in the
implemented solution is that any measurement of data is considered to be
persisted without further analysis.

All in all, the DSP Data Persistence component implemented for the NetBEAMS
solution to SF-BEAMS can be used in any network server. The solution for data
persistence is a novice way to save data into the database, provided by
mongoDB's ability to cope with the uncertainty of any properties collected from
any sensor specification. Furthermore, the solution also provides a good
support to the provenance-related data.
